---
title: "Data Science for Public Policy"
subtitle: "Final Project"
author: "Asad Azhar, Juan Menendez, Priscila Stisman, Talia Stringfellow"
execute:
  warning: false
format:
  html:
    embed-resources: true
---
## Final Project Studying Latin America

# Packages
```{r}
library(tidyverse)
library(tidymodels)
library(recipes)
library(ggplot2)
library(patchwork)
library(tidyclust)
library(dplyr)
library(themis)
library(haven)
library(rpart)
library(lubridate)
library(archive)
library(utils)
library(tigris)
library(sf)
library(geodata)
library(rnaturalearth)
library(mapview)
library(textrecipes) 
library(vip)

library(rsample)
library(recipes)
library(parsnip)
library(workflows)
library(tune)
library(yardstick)
```

# Read in Data

```{r}
lat_am_pop_2023 <- read_dta("data/Merge_2023_LAPOP_AmericasBarometer_v1.0_w.dta")

```

## Data cleaning

In this section, we identify and clean our variables of intrest

Variables of interest:

[**Etnicity/ skin color identification**]{.underline}

Colorr: contains individual skintone on a scale form 0 to 11 (categorical)

Colori: interviewer skincolor

Etid: self perceived ethnicity (categorical)

[**Employment Status**]{.underline}

Ocup4a: employment status (dummy)

formal: formal job (dummy)

[**Social Capital**]{.underline} 

cp13: political party meetings attendance (frequency of attendance, categorical)

cp8: community meetings attendance (frequency of attendance, categorical)

r4a: has a mobile phone (dummy)

smedia1n: has social media (dummy)

[**Demographic Data**]{.underline}

q1tc_r: sex (make dummy)

q2: age (continuous)

q12cn: number of members in the household

q10inc: income (Argentine pesos, categorical) (for Argentina)

Edre: education attainment (all levels)

pais : country (numeric)

```{r}

Workingdata <- lat_am_pop_2023 %>% select(colorr, colori, etid, ocup4a, formal, cp13, cp8, r4a, smedia1n, q1tc_r, q2, q12cn, q10inc, edre, pais, ur, q14motan)

Workingdata <- Workingdata %>%
  mutate(ocup4a = case_when(
    ocup4a == 1 ~ 1,
    ocup4a == 2 ~ 0,
    ocup4a == 3 ~ 0,
    ocup4a == 4 ~ 0,
    ocup4a == 5 ~ 0,
    ocup4a == 6 ~ 0,
    ocup4a == 7 ~ 0,
  ))

Workingdata <- Workingdata %>%
  mutate(cp13 = case_when(
    cp13 == 4 ~ 1,
    cp13 == 3 ~ 2,
    cp13 == 2 ~ 3,
    cp13 == 1 ~ 4,
  ))

Workingdata <- Workingdata %>%
  mutate(cp8 = case_when(
    cp8 == 4 ~ 1,
    cp8 == 3 ~ 2,
    cp8 == 2 ~ 3,
    cp8 == 1 ~ 4,
  ))

Workingdata <- Workingdata %>%
  mutate(q1tc_r = case_when(
    q1tc_r == 1 ~ 1,
    q1tc_r == 2 ~ 0,
    cp8 == 3 ~ 1,
  ))

#cleaning ethnicity 
Workingdata <- Workingdata %>%
  mutate(etid = case_when(
    etid == 1 ~ 1, 
    etid == 2 ~ 2, 
    etid == 3 ~ 3, 
    etid == 4 ~ 4, 
    etid == 5 ~ 5, 
    etid == 6 ~ 6, 
    etid == 7 ~ 7, 
    etid == 1506 ~ NA,
    etid == 1106 ~ NA,
    etid == 1110 ~ NA, 
    TRUE ~ NA
  ))

# create dummy variable for college
Workingdata <- Workingdata %>%
  mutate(university = case_when(
    edre == 6 ~ 1,
    edre == 5 ~ 1,
    !(edre %in% c(5, 6)) ~ 0,
    TRUE ~ NA
  ))

Workingdata <- Workingdata %>%
  mutate(university_factor = factor(case_when(
    edre == 6 ~ "University",
    edre == 5 ~ "University", 
    !(edre %in% c(5, 6)) ~ "Non-University", 
    TRUE ~ NA_character_ 
  )))

skincolor <- c("#fff5f6", "#f5e2dc", "#e9c1b7", "#e7c9a5", "#c0a280", "#9d7c53", "#85674f", "#70503b", "#523c2f", "#422811", "#383127")

skincodes <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11)
skincolorpalette <- tibble(skincodes, skincolor)

educ_labels <- c(
  "None" = 0,
  "Primary School Incomplete" = 1,
  "Primary School Complete" = 2,
  "High-school Incomplete" = 3,
  "High-school Complete" = 4,
  "Tertiary/University Incomplete" = 5,
  "Tertiary/University Complete" = 6
)


#cleaning Income 
income_codes <- c(1701, 1702, 1703, 1704, 1705, 1706, 1707, 1708, 1709, 1710, 1711, 1712, 1713, 1714, 1715)

# Labels
income_labels <- c("Entre $0 y $14.000 pesos", "Entre $14.001 y $26.000 pesos", "Entre $26.001 y $39.000 pesos",
            "Entre $39.001 y $47.000 pesos", "Entre $47.001 y $55.000 pesos", "Entre $55.001 y $61.000 pesos",
            "Entre $61.001 y $68.000 pesos", "Entre $68.001 y $81.000 pesos", "Entre $81.001 y $88.000 pesos",
            "Entre $88.001 y $97.000 pesos", "Entre $97.001 y $115.000 pesos", "Entre $115.001 y $136.000 pesos",
            "Entre $136.001 y $168.000 pesos", "Entre $168.001 y $221.000 pesos", "Más de $221.000 pesos")

# Assign labels
income_values <- factor(income_codes, levels = income_codes, labels = income_labels)

#creating dummy for stating Lack of educational opportunities as reason for why they have considered emigrating 
Workingdata <- Workingdata %>%
  mutate(leave4educ = case_when(
    q14motan == 3 ~ 1,
    !(q14motan %in% 3) ~ 0,
    TRUE ~ NA
  ))

#Adding country labels 
Workingdata <- Workingdata %>% 
  filter(pais < 39) %>% #getting rid of US and Canada from dataset 
  mutate(country_name = factor(case_when(
    pais == 1 ~ "Mexico",
    pais == 2 ~ "Guatemala", 
    pais == 3 ~ "El Salvador",
    pais == 4 ~ "Honduras", 
    pais == 5 ~ "Nicaragua",
    pais == 6 ~ "Costa Rica", 
    pais == 7 ~ "Panama",
    pais == 8 ~ "Colombia", 
    pais == 9 ~ "Ecuador",
    pais == 10 ~ "Bolivia", 
    pais == 11 ~ "Peru", 
    pais == 12 ~ "Paraguay",
    pais == 13 ~ "Chile", 
    pais == 14 ~ "Uruguay",
    pais == 15 ~ "Brazil", 
    pais == 16 ~ "Venezuela",
    pais == 17 ~ "Argentina", 
    pais == 21 ~ "Dominican Republic", 
    pais == 22 ~ "Haiti", 
    pais == 23 ~ "Jamaica",
    pais == 24 ~ "Guyana", 
    pais == 25 ~ "Trinidad & Tobago",     
    pais == 26 ~ "Belize",
    pais == 27 ~ "Suriname", 
    pais == 28 ~ "Bahamas", 
    pais == 29 ~ "Barbados",
    pais == 30 ~ "Grenada", 
    pais == 31 ~ "Saint Lucia", 
    pais == 32 ~ "Dominica", 
    pais == 33 ~ "Antigua and Barbuda", 
    pais == 34 ~ "Saint Vincent and the Grenadines", 
    pais == 35 ~ "Saint Kitts and Nevis",
    TRUE ~ NA_character_ 
  ))) %>%
  relocate(country_name, .after = pais) 

SouthAmerica_only <- Workingdata %>% 
  filter(country_name == "Argentina" 
         | country_name == "Uruguay" 
         | country_name == "Chile" 
         | country_name == "Paraguay" 
         | country_name == "Bolivia" 
         | country_name == "Peru" 
         | country_name == "Ecuador" 
         | country_name == "Brazil" 
         | country_name == "Suriname" 
         | country_name == "Guyana" 
         | country_name == "French Guiana"  
         | country_name == "Venezuela" 
         | country_name == "Colombia")

SouthAmerica_only <- SouthAmerica_only %>%
  rename(skin_color = colorr,
         ethnicity = etid,
         employed = ocup4a,
         mobile = r4a, 
         social_media = smedia1n,
         gender = q1tc_r,
         age = q2, 
         income = q10inc,
         educ_attainment = edre,
         political_involvement = cp13,
         community_involvement  =cp8, 
         country_num = pais)

```


